---
title: "Data preprocessing"
author: "Sergio Picart-Armada"
date: "December 28, 2017"
output: html_document
---

```{r setup, include=FALSE}
# Show the code, but suppress messages and warnings
# otherwise document gets too long
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Getting started

```{r}
library(STRINGdb)
library(igraph)
library(ggplot2)
library(diffuStats)

library(plyr)
library(dplyr)
library(tidyr)

# have all config variables in a different env
config <- new.env(parent = globalenv())
source("20_config.R", local = config)

# abbreviations
source(config$file_abbrev)
stopifnot(exists("disease_abbrev"))
```

# Build STRING network

## Read STRING and OpenTargets data

```{r}
# read tabular file with gene-disease data
# IMPORTANT: keep quote = "" to disable quotation, as
# gene names can have quotes that malform the object
df_disease <- read.table(
  file = config$file_alldiseases, header = T, sep = "\t", 
  stringsAsFactors = FALSE, quote = "") 
df_disease$known_gene_binary <- (df_disease[[config$col_genetic]] >= config$threshold_genetic)*1

# histogram of the genetic associations (justifying threshold)
v.hist <- df_disease[[config$col_genetic]] %>% subset(. > 0)
gg.hist <- ggplot(
  data.frame(genetic = v.hist), 
  aes(x = genetic)) + 
  geom_histogram(binwidth = .02) + 
  geom_vline(xintercept = config$threshold_genetic, lty = 2) + 
  theme_bw() + 
  theme(aspect.ratio = 1) + 
  xlab("Genetic association score from OpenTargets") + 
  ylab("Number of non-zero associations") + 
  ggtitle("Cutoff for higher confidence genetic associations", 
          subtitle = paste0("Threshold: score >= ", config$threshold_genetic))
ggsave(filename = paste0(config$dir_data, "/descriptive_genetic_scores.svg"), 
       device = NULL, plot = gg.hist)
ggsave(filename = paste0(config$dir_data, "/descriptive_genetic_scores.eps"), 
       device = NULL, plot = gg.hist)

# summary as tex file
sink("/dev/null")
stargazer::stargazer(
  df_disease, nobs = TRUE, mean.sd = TRUE, median = TRUE, 
  iqr = TRUE, out = paste0(config$dir_data, "/descriptive_rawdata.tex"))
sink()

paste("Number of NAs:", sum(is.na(df_disease)))

# retrieve string db
string_db <- do.call(
  STRINGdb$new, config$params_string
)

# summary in text format of the string version
writeLines(
  capture.output(print(config$params_string)), 
  con = paste0(config$dir_data, "/info_string.txt")
)


# network object
string_g <- string_db$get_graph()
summary(string_g)
```

## Map genes from OpenTargets to ENSEMBL

```{r}
# Map genes to STRING
df_map <- string_db$map(
  df_disease, 
  my_data_frame_id_col_names = config$colname_symbol)
rows_original <- nrow(df_map)

# we only lose 3% of the ids
df_map <- na.omit(df_map)
rows_mapped <- nrow(df_map)

paste("Number of rows after mapping:", rows_mapped)
paste("Fraction of data rows correctly mapped:", 
      rows_mapped/rows_original)


# are ensembl ids unique, given a disease?
col_split <- c("STRING_id", "disease.id")
df_collisions <- plyr::ddply(df_map, col_split, function(df) if (nrow(df) > 1) df)

# View genes that collide
# In the last run, this data frame was empty: no collisions
df_collisions

# fix collisions, if any
if (nrow(df_collisions)) {
  df_map <- plyr::ddply(
    df_map, col_split, 
    function(df) {
      n <- nrow(df)
      if (n == 1) return(df)
      ord <- order(
        df$known_drug_binary, # pick genes with drugs in case of ties
        runif(n), # break ties at random
        decreasing = TRUE)
      df[ord[1], , drop = FALSE]
    }, 
    .progress = "text"
  )
}
rows_final <- nrow(df_map)

paste("Final dimensions:", paste(dim(df_map), collapse = " by "))

# save in a text file 
writeLines(
  paste(
    c("Number of rows"), 
    c("in the original data:", "rows mapping to STRING:", "rows when removing duplicates:"), 
    c(rows_original, rows_mapped, rows_final), 
    sep = " ", collapse = "\n"
  ), 
  con = paste0(config$dir_data, "/descriptive_mapping.txt")
)
```

## Filter network and assess coverage

```{r}
# same filters as in takeda
# These filters specify which edges will drop
list.edge.filters <- list(
  Net1 = "combined_score < 400 | experiments < 1", 
  Net2 = "experiments < 600", 
  Net3 = "experiments < 400 & database < 400", 
  Net4 = "combined_score < 700 | (experiments < 1 & database < 1)", 
  Net5 = "combined_score < 700 | (experiments < 1 & database < 1 & textmining < 900)", 
  Net6 = "database < 400", 
  All = "combined_score < 0"
)

# Build all the networks, take their largest connected component 
# and count how many genes actually map to them
df_coverage <- plyr::ldply(
  list.edge.filters, 
  function(quote.char) {
    # network 
    g <- delete_edges(
      string_g, 
      E(string_g)[eval(parse(text = quote.char), envir = environment())]
    )
    
    # quantify what happens before and after taking largest cc
    plyr::ldply(
      list(before_cc = identity, after_cc = largest_cc), 
      function(fun) {
        g <- do.call(fun, list(g))
        
        # number of rows of mapped data that also map to the graphs
        rows_mapped <- df_map$STRING_id %in% V(g)$name
        data.frame(
          filter = quote.char, 
          nodes = vcount(g), 
          edges = ecount(g), 
          coverage_allrows = sum(rows_mapped), 
          coverage_drug = sum(df_map$known_drug_binary*rows_mapped), 
          coverage_genetic = sum(df_map$known_gene_binary*rows_mapped), 
          stringsAsFactors = FALSE
        )
      }, 
      .id = "cc"
    )
  }, 
  .id = "network", 
  .progress = "text"
)

write.csv(df_coverage, 
          file = paste0(config$dir_data, "/networks_coverage.csv"), 
          row.names = FALSE)

df_coverage

# this assumes the cc level is alphabetically sorted
tex_coverage <- df_coverage %>% 
  group_by(network, filter) %>% 
  summarise_all(function(x) paste0(x[2], "(", x[1], ")")) %>% 
  select(-cc) %>% 
  xtable::xtable(x = .)

xtable::print.xtable(
  tex_coverage, 
  include.rownames = FALSE, 
  file = paste0(config$dir_data, "/descriptive_networks.tex"))
```

```{r}
# Plot the coverage to decide
df_plotcoverage <- tidyr::gather(
  filter(df_coverage, cc == "after_cc"), key = "coverage_type", value = coverage_value, 
  contains("coverage"))
ggplot(df_plotcoverage, aes(x = nodes, y = coverage_value)) + 
  geom_smooth(method = "lm") + 
  geom_text(aes(label = network), vjust = "inward") + 
  facet_wrap(~coverage_type, scales = "free_y") + 
  scale_x_continuous(expand = c(.15, .15)) + 
  theme_bw() + 
  theme(aspect.ratio = 1) + 
  ggtitle("Coverage of input genes in the network")
ggsave(filename = paste0(config$dir_data, "/networks_coverage.png"), 
       width = 9, height = 4)
```

# Save network

## Choose filter based on coverage

```{r}
# Pick network 4: equilibrium between mapped genes and simplicity
choice_net <- "Net4"
choice_filter <- list.edge.filters[choice_net]

# delete edges not meeting the filter
g_filter <- delete_edges(
  string_g, 
  E(string_g)[eval(parse(text = choice_filter), 
                   envir = environment())]
  ) %>% largest_cc

# Edge weights should lie between 0 and 1
E(g_filter)$weight <- E(g_filter)$combined_score/max(E(g_filter)$combined_score)

# Save metadata to the graph object
g_filter$choice_filter <- choice_filter
g_filter$config <- config

# final graph
summary(g_filter)
stopifnot(is.simple(g_filter))
stopifnot(is.connected(g_filter))

# delete the edge attributes other than the weight
# to save space
attr_edge <- list.edge.attributes(g_filter)
for (att in setdiff(attr_edge, "weight")) {
  g_filter <- delete_edge_attr(g_filter, att)
}
```

## Clean OpenTargets dataset

```{r}
# add the dataset that maps to the graph
df_dataset <- filter(df_map, STRING_id %in% V(g_filter)$name)

col_disease <- "disease.efo_info.label"

df_count <- plyr::ddply(
  df_dataset, 
  col_disease, 
  function(df) {
    data.frame(
      drug = sum(df$known_drug_binary), 
      genetic = sum(df$known_gene_binary)
    )
  })
write.csv(
  df_count, 
  file = paste0(config$dir_data, "/diseases_mapped_genes.csv"), 
  row.names = FALSE)

# export latex table
xtable::print.xtable(
  xtable::xtable(df_count, digits = 0), 
  include.rownames = FALSE, 
  file = paste0(config$dir_data, "/descriptive_mapped_genes.tex"))
```

## Filter diseases by a minimum of drugs and genetic genes

```{r}
# range to sweep
thresholds_nmin <- 1:50
# numbes to output disease lists 
thresholds_txt <- c(5, 10, 20, 50)

df_filterdisease <- plyr::ldply(
  thresholds_nmin, 
  function(threshold) {
    df_filt <- filter(df_count, drug >= threshold & genetic >= threshold)
    df <- data.frame(
      threshold = threshold, 
      diseases = nrow(df_filt)
    )
    # write diseases passing the filter
    if (threshold %in% thresholds_txt) {
      writeLines(
        df_filt$disease.efo_info.label, 
        con = paste0(config$dir_data, "/diseases_over_", threshold, "_genes.txt")
      )
    }
    df
  }
)
ggplot(df_filterdisease, aes(x = threshold, y = diseases)) + 
  geom_bar(stat = "identity") + 
  xlab("Minimum associated genes") +
  ylab("Diseases above threshold") + 
  theme_bw() + 
  theme(aspect.ratio = 1) + 
  ggtitle("Number of diseases after imposing minimum genes", 
          subtitle = "Threshold applies to both drugs and genetic associations")
ggsave(filename = paste0(config$dir_data, "/diseases_thresholds_ngenes.png"), 
       width = 5, height = 5)
```

## Load and map complex data

```{r}
# rda file from David
load(config$file_complexes)

# the list in David's file is called "a" and we want the gene-complex mapping
# it is stored in a$all$genes.for.complexes
# (current criteria is in favour of using "all")
list_complex_raw <- a$all$genes.for.complexes

# descriptive stats
df_descriptive_complex <- plyr::ldply(
  list_complex_raw, 
  function(prot) {
    data.frame(
      original = length(unique(prot)), 
      mapped = sum(paste0("9606.", as.character(prot)) %in% V(g_filter)$name)
    )
  }, 
  .id = "complex_id"
)

# insert NAs instead of empty complexes
df_descriptive_complex[df_descriptive_complex == 0] <- NA

sink("/dev/null")
stargazer::stargazer(
  df_descriptive_complex, nobs = TRUE, mean.sd = TRUE, median = TRUE, 
  iqr = TRUE, out = paste0(config$dir_data, "/descriptive_complexes.tex"))
sink()

# complex data as a list
# genes are mapped to the network, only the intersection is kept
list_complex <- plyr::llply(
  list_complex_raw, 
  function(prot) {
    intersect(paste0("9606.", as.character(prot)), V(g_filter)$name)
  } 
)

# as a data frame
df_complex <- plyr::ldply(
  list_complex, 
  function(prot) data.frame(STRING_id = prot), 
  .id = "complex_id"
)
```

## Save the dataset and the network

```{r}
# filter diseases
disease_filtered <- filter(
  df_count, drug >= config$threshold_ngenes & 
  genetic >= config$threshold_ngenes)$disease.efo_info.label

# add OT and complex data to the network
g_filter$dataset <- filter(df_dataset, disease.efo_info.label %in% disease_filtered)
g_filter$list_complex <- list_complex
g_filter$df_complex <- df_complex

# save the graph with the dataset
save(g_filter, file = config$graph_alldiseases, compress = "xz")
```

# Input data

## Input data 

```{r}
# the dataset from OpenTargets
df_disease_formatted <- mutate(
  g_filter$dataset, 
  disease = as.factor(disease.efo_info.label)) %>% 
  rename(drugs = known_drug_binary, genetic = known_gene_binary) %>% 
  mutate(validation = drugs)

n_total <- vcount(g_filter)
```

```{r}
# data frame with input scores in long format
# use drugs and genetic as input streams
df_input <- plyr::ddply(
  reshape2::melt(
    df_disease_formatted, measure.vars = c("drugs", "genetic"), 
    variable.name = "input_type", value.name = "input"),
  c("disease", "input_type"), 
  function(disease) {
    # browser()
    nm <- V(g_filter)$name
    x <- nm %in% (filter(disease, input == 1)$STRING_id)
    val <- nm %in% (filter(disease, validation == 1)$STRING_id)
    
    data.frame(
      STRING_id = nm, 
      score = x*1, 
      validation = val*1
    )
  }, 
  .progress = "text"
)

# Save the drugs data as a long data frame
df_drugs <- filter(df_input, input_type == "drugs") %>%
  dplyr::select(disease, STRING_id, validation) %>% 
  rename(drugs = validation)
# pairs disease-gene (drugs == 0 not included to save space)
write.csv(filter(df_drugs, drugs == 1L) %>% dplyr::select(-drugs), 
          file = paste0(config$dir_data, "/data_disease_drugs.csv"), 
          row.names = FALSE)

# count in how many diseases genes participate
# Only write genes in many diseases
df_ndisease <- plyr::ddply(
  df_drugs, "STRING_id", 
  function(df) c(n_disease = sum(df$drugs))) %>% 
  dplyr::filter(n_disease >= 10) %>%
  dplyr::arrange(desc(n_disease))
write.csv(df_ndisease, 
          file = paste0(config$dir_data, "/descriptive_disease_count_genes.csv"), 
          row.names = FALSE)

# Small tests
local({
  a <- subset(df_input, input_type == "drugs")
  b <- subset(df_input, input_type == "genetic")
  tr <- all(a$score == a$validation) & all(a$score == b$validation)  
  stopifnot(tr)
})

# select streams other than drug and overall
# to use them as "reference streams" in the benchmark
# data frame in long format
df_streams <- dplyr::select(
  df_disease_formatted, disease, STRING_id, 
  contains("association_score"), -contains("drug"), 
  -contains("overall")) %>% 
  gather(key = stream, value = score, 
         contains("association_score"), factor_key = TRUE) %>%
  ddply(
    c("disease", "stream"), 
    function(df_dis) {
      nm <- V(g_filter)$name
      x <- nm %in% df_dis$STRING_id
      
      original <- setNames(numeric(length(nm)), nm)
      original[as.character(df_dis$STRING_id)] <- df_dis$score
      
      data.frame(STRING_id = nm, score = original)
    }, 
    .progress = "text"
)
```

```{r}
# Descriptive data about the diseases: 
# number of genetic genes, number of drugs genes, 
# overlap (with p-value and fdr)
df_descriptive <- plyr::ddply(
  dplyr::select(df_input, -validation) %>% 
    reshape2::dcast(STRING_id+disease~input_type, 
                    fun.aggregate = NULL, value.var = "score"), 
  "disease", 
  function(df) {
    tibble(
      n_genetic = sum(df$genetic), 
      n_drug = sum(df$drugs), 
      overlap = sum(df$genetic*df$drugs), 
      p_value = fisher.test(x = df$genetic, y = df$drugs)$p.value
    ) 
  }, 
  .id = "disease") %>% 
  mutate(fdr = p.adjust(p_value, method = "fdr"))

write.csv(df_descriptive, 
          file = paste0(config$dir_data, "/descriptive_diseases.csv"), 
          row.names = FALSE)

# export in latex format
xtable::xtable(
  df_descriptive, 
  digits = c(0, 0, 0, 0, 0, 2, 2), 
  display = c("s", "d", "d", "d", "d", "e", "e")) %>% 
  xtable::print.xtable(x = ., include.rownames = FALSE, 
                       file = paste0(config$dir_data, "/descriptive_diseases.tex"))

# check for NAs in the input
stopifnot(all(!is.na(df_input)))

# sanity check: all diseases must contain 1's but have a mean closer to 0
# (most genes are negatives)
group_by(df_input, disease, input_type) %>%
  dplyr::select(score) %>%
  summarise_all(c("min", "mean", "max"))
```

# Complex-aware cross validation

## Precompute cross validation folds

```{r}
set.seed(1)

# list with validation folds
# the format is list$disease$cv_scheme${train,validation}
# train and validation contain the indices of the genes that 
# should be used for both purposes. 
# If validation is NULL, then the complementary of train should be used
list_cv_folds <- plyr::dlply(
  # filter(df_input, disease %in% c("allergy", "chronic obstructive pulmonary disease")), 
  filter(df_input, input_type == "drugs"),
  "disease", 
  function(df_in) {
    # browser()
    ans <- list()
    
    name_disease <- as.character(df_in$disease[1])

    # take response only
    y <- setNames(df_in$validation, df_in$STRING_id)
    
    #######################################
    # First approach: classic stratified CV
    folds_classic <- caret::createMultiFolds(
      y = as.factor(y), 
      k = config$k_cv, 
      times = config$times_cv)
    # Rename to be consistent with other schemes
    names_classic <- names(folds_classic) %>% 
      strsplit(split = "\\.") %>%  
      sapply(function(x) paste(rev(x), collapse = "."))  
    names(folds_classic) <- names_classic
    
    ans$classic <- lapply(
      folds_classic, 
      function(x) list(train = x)
    )
    
    #######################################
    # Complex-aware approaches
     
    # First step: find complexes relevant to the disease
    # gene names
    gene_disease <- names(y)[y == 1L] %>% unique
    # find their associated complexes
    df_compl <- filter(df_complex, STRING_id %in% gene_disease)
    # genes within the disease that belong to complexes
    gene_compl <- df_compl$STRING_id %>% unique
    # identifiers of the complexes related to the disease
    id_compl <- df_compl$complex_id %>% as.character %>% unique
    # have the complexes as a list
    list_whole_compl <- list_complex[id_compl]
    gene_whole_compl <- list_whole_compl %>% unlist %>% unique
    
    # Second step: generate graph with genes as nodes for merging complexes
    # 
    # union of disease genes and complex genes for that disease
    # (because some complex genes might not be within the disease - 
    # this should not happen too often though)
    gene_union <- union(gene_whole_compl, gene_disease)
    n <- length(gene_union)
    # First, an empty graph
    g <- graph.empty(n)
    V(g)$name <- gene_union
    # Modify its adjacency matrix by adding edges between all the 
    # genes in a complex
    mat_adj <- get.adjacency(g)
    for (com in list_whole_compl) {
      mat_adj[com, com] <- 1
    }
    # Redefine the graph with the right adjacency
    g <- graph_from_adjacency_matrix(mat_adj, mode = "undirected")
    g_components <- clusters(g)
    
    # Get the labels for each gene
    labels_gene <- g_components$membership
    labels <- 1:(g_components$no)
    # Number of genes in each component
    labels_size <- g_components$csize
    # Compute the number of disease genes per component
    # (because some of them might not be fully made of disease genes...)
    labels_positives <- sapply(
      labels, 
      function(lab) {
        genes <- names(labels_gene)[labels_gene == lab]
        length(intersect(genes, gene_disease))
      }
    )
    stopifnot(all(labels_size >= labels_positives))
    
    # ids of the rest of vertices
    genes_outside <- V(g_filter)[!(name %in% V(g)$name)] %>% as.numeric
    
    # names for repetitions (with trailing 0s if needed)
    names_rep <- paste0(
      "Rep", 
      formatC(1:config$times_cv, 
              width = ceiling(log10(config$times_cv)), 
              format = "d", flag = "0"))
    #######################################
    # Complex-aware 1st method: block CV
    # browser()
    # all the splits in a list
    # Format: list$RepX$FoldY and list$RepX.FoldY after unlisting
    ans$block <- plyr::llply(
      setNames(names_rep, names_rep), 
      function(rep_number) {
        # browser()
        # shuffle the components
        shuffle_labels <- sample(labels)
        # cumulative number of disease genes in this ordering
        shuffle_sum <- labels_positives[shuffle_labels]
        shuffle_cumsum <- cumsum(shuffle_sum)
        
        # Total number of disease genes
        total_sum <- length(gene_disease)
        stopifnot(total_sum == tail(shuffle_cumsum, 1))
        
        # Theoretical endpoints
        endpoints <- ((1:config$k_cv - 1)/config$k_cv)*total_sum
        # Find the closest cuts
        cuts <- sapply(endpoints, function(x) which.min(abs(shuffle_cumsum - x)))
        # In which cut is each position?
        # This leaves the positives (and maybe some negatives) with a defined fold
        folds_inside <- sapply(seq_along(shuffle_sum), function(x) sum(x >= cuts)) %>% 
          paste0("Fold", .) %>% 
          split(x = shuffle_labels, f = .) %>% 
          llply(function(labels_fold) {
            # get the ids of the nodes in each fold
            genes_fold <- V(g)[labels_gene %in% labels_fold]$name
            V(g_filter)[name %in% genes_fold] %>% as.numeric
          })
        
        # Assuming that the number of negatives in the complexes is small.. 
        # the remaining negatives will be equally split among folds
        folds_outside <- caret::createFolds(genes_outside, k = config$k_cv) %>% 
          llply(function(ids_fold) {
            genes_outside[ids_fold]
          })

        # Merge both lists and take the complementary, 
        # as each fold contains the training data.
        lapply(
          setNames(names(folds_inside), names(folds_inside)), 
          function(fold) {
            list(train = setdiff(
              1:vcount(g_filter), 
              c(folds_inside[[fold]], folds_outside[[fold]])
              )
            )
          }
        )
      }
    ) %>% unlist(recursive = FALSE, use.names = TRUE)
    
    #######################################
    # Complex-aware 2nd method: pick one representative per block
    ans$representative <- plyr::llply(
      setNames(names_rep, names_rep), 
      function(rep_number) {
        # for each merged complex, pick one representative
        # the rest of genes belong to "excluded"
        
        # representatives (by string id)
        # This will change between repetitions!
        list_repr <- lapply(
          labels, 
          function(lab) {
            # find genes and permute them
            genes <- names(labels_gene)[labels_gene == lab]
            genes_disease <- intersect(genes, gene_disease) %>% sample
            
            # pick the first gene as representative and the rest as excluded
            genes_disease[1]
          }
        ) %>% unlist
        
        # new vector of labels
        y_new <- setNames(V(g_filter)$name %in% list_repr, names(V(g_filter)))*1L
        
        # list of excluded genes
        list_excluded <- setdiff(V(g)$name, list_repr)
        id_excluded <- which(V(g_filter)$name %in% list_excluded)
        
        # browser()
        # stratified partition of the new training vector
        folds <- caret::createFolds(as.factor(y_new), 
                                    k = config$k_cv, 
                                    returnTrain = TRUE) 
        
        # return 
        list_folds <- lapply(
          setNames(names(folds), names(folds)), 
          function(fold) {
            list(
              train = setdiff(folds[[fold]], id_excluded), 
              validation = setdiff(1:n_total, c(folds[[fold]], id_excluded))
            )
          }
        )
        
        list_folds
      }
    ) %>% unlist(recursive = FALSE, use.names = TRUE)
    
    ans
  }, 
  .progress = "text"
)
```


## Sanity check of folds


```{r}
df_stat_complex <- plyr::ldply(
  setNames(names(list_cv_folds), names(list_cv_folds)), 
  function(disease_name) {
    # binary drugs data
    df_dis <- filter(df_input, disease == disease_name & 
                       input_type == "drugs" & score == 1L)
    vec_dis <- setNames(
      V(g_filter)$name %in% df_dis$STRING_id, 
      V(g_filter)$name)
    gene_dis <- names(vec_dis)[vec_dis == 1L]
    
    df_complex_in_disease <- filter(df_complex, STRING_id %in% gene_dis)
    
    # check all the cross validation schemes
    plyr::ldply(
      list_cv_folds[[disease_name]], 
      function(scheme) {
        # and all the folds within it
        plyr::ldply(
          scheme, 
          function(rep_fold) {
            # browser()
            
            id_train <- unique(rep_fold$train)
            id_val <- unique(rep_fold$validation)
            if(is.null(id_val)) id_val <- -id_train
            
            vec_train <- vec_dis[id_train]
            vec_val <- vec_dis[id_val]
            
            # breaks any complex?
            # see if all the complex genes involved with the disease
            # are inside or outside
            # the training set
            # i.e. either 0% or 100% of their genes map to it within 
            # the training set
            complex_mapped <- plyr::daply(
              df_complex_in_disease, 
              "complex_id", 
              function(df_comp) {
                # browser()
                genes_comp <- df_comp$STRING_id %>% as.character
                in_train <- any(genes_comp %in% names(vec_train))
                in_val <- any(genes_comp %in% names(vec_val))
                # complexes are split if and only if we have 
                # positives on training and on validation
                in_train & in_val
              }
            ) 
            # NAs are complexes that have no members 
            complex_split <- sum(complex_mapped, na.rm = TRUE)
            
            data.frame(
              train_all = length(vec_train), 
              train_pos = sum(vec_train), 
              validation_all = length(vec_val), 
              validation_pos = sum(vec_val), 
              split_complexes = complex_split
            )
          }, 
          .id = "fold"
        )
      }, 
      .id = "cv_scheme"
    )
  }, 
  .id = "disease", 
  .progress = "text"
)

# make sure there are always 
stopifnot(all(df_stat_complex$train_pos > 0))
stopifnot(all(df_stat_complex$train_val > 0))

# we can see that block cv discarded some folds
# due to empty train or val
table(df_stat_complex$cv_scheme)
summary(df_stat_complex)

# generate tex tables
# summary of the npositives and split complexes
df_stat_cvschemes <- dplyr::group_by(df_stat_complex, disease, cv_scheme) %>% 
  summarise_if(
    is.numeric, 
    function(x) {
      paste0(
        format(mean(x), digits = 0, nsmall = 1), "(", 
        format(sd(x), digits = 0, nsmall = 2), ")"
      )
    }) %>% 
  select(disease, cv_scheme, train_pos, validation_pos, split_complexes) %>% 
  reshape2::melt(id.vars = c("disease", "cv_scheme")) %>% 
  reshape2::dcast(disease ~ variable + cv_scheme) 
tex_stat_cvschemes <- knitr::kable(
  df_stat_cvschemes, 
  format = "latex", booktabs = TRUE, escape = TRUE) %>% 
  kableExtra::kable_styling(latex_options = c("striped", "scale_down")) %>% 
  kableExtra::add_header_above(c(" " = 1, "train_pos" = 3, "validation_pos" = 3, "split_complexes" = 3))
writeLines(tex_stat_cvschemes, con = paste0(config$dir_data, "/descriptive_cvsplits.tex"))

tex_count_cvschemes <- knitr::kable(
  dplyr::group_by(df_stat_complex, cv_scheme) %>% summarise(count = length(train_pos)), 
  format = "latex", booktabs = TRUE, escape = TRUE) %>% 
  kableExtra::kable_styling(latex_options = c("striped", "scale_down"))
writeLines(tex_count_cvschemes, con = paste0(config$dir_data, "/descriptive_cvsplits_counts.tex"))

# check that "classic" and "representative" are balanced
# we substract the maximum and the minimum amount of genes 
# in training and validation
# If this is 1 or less, it is balanced. 
# If not,
df_check <- dplyr::group_by(df_stat_complex, disease, cv_scheme) %>% 
  summarise(range_train = max(train_pos) - min(train_pos), 
            range_val = max(validation_pos) - min(validation_pos))

# Check balance (meaning that cv is stratified as expected)
stopifnot(all(filter(df_check, cv_scheme != "block")$range_train <= 1))
stopifnot(all(filter(df_check, cv_scheme != "block")$range_val <= 1))

# check number of split complexes by cv scheme
dplyr::group_by(df_stat_complex, cv_scheme) %>% 
  summarise(split_complexes_total = sum(split_complexes))


dplyr::group_by(df_stat_complex, disease, cv_scheme) %>% 
  summarise(npos_mean = mean(train_pos), 
            npos_sd = sd(train_pos))
```

```{r}
# plot the data imbalance within splits
df_plot_cvsplits <- mutate(
  df_stat_complex, 
  proportion = validation_pos/(train_pos + validation_pos), 
  disease = factor(disease_abbrev[as.character(disease)], levels = disease_abbrev)) 
palette25 <- readLines(config$file_palette25)

gg <- ggplot(df_plot_cvsplits, aes(x = disease, y = proportion, fill = disease)) + 
  geom_hline(yintercept = 1/config$k_cv, lwd = .5, color = "gray50") +
  geom_boxplot(outlier.size = .05, lwd = .2) + 
  scale_fill_manual(values = palette25, guide = FALSE) + 
  facet_wrap(~cv_scheme) + 
  coord_flip() + 
  theme_bw() + 
  xlab("") + 
  ylab("Proportion of drugs-related genes in the validation fold") + 
  theme(aspect.ratio = 3, 
            axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 6.5), 
            legend.position = "none")

ggsave(filename = paste0(config$dir_data, "/descriptive_folds_imbalance.svg"), 
       device = NULL, plot = gg)
ggsave(filename = paste0(config$dir_data, "/descriptive_folds_imbalance.eps"), 
       device = NULL, plot = gg)
```


```{r}
# Save CV schemes and input data frames
save(df_input, df_streams, list_cv_folds, 
     file = config$file_input, 
     compress = "xz")

write.csv(df_stat_complex, 
          file = paste0(config$dir_data, "/descriptive_cv_splits.csv"), 
          row.names = FALSE)
```

# Compute and save kernel (computationally intensive)

```{r}
if (!file.exists(config$file_kernel)) {
  # Compute kernel
  # last time: ~1000s
  kernel_time <- system.time({
    K <- diffuStats::regularisedLaplacianKernel(
      graph = g_filter, 
      normalized = config$kernel_normalised
    )
  })
  kernel_time
  
  # Standard compression; tried xz but it is not 
  # worth it and takes too long
  save(K, file = config$file_kernel)  
}
```

# Reproducibility

```{r}
out <- capture.output(sessionInfo())
writeLines(out, con = paste0(config$dir_metadata, "/20_sessionInfo_preprocessing.txt"))
```

